Characterizing Optimizations to Memory Access Patterns using Architecture-Independent Program Features

### 2020-03-21 paper/paper.tex

	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelizable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
	%The language provides support for individual \emph{kernels}, portions of parallelizable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
	This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.

@@

	The code is further optimized by improving the locality of reads from Matrix $B$.
	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
	The repeated reads of elements of matrix $B$ can be shared between threads in a work-group by reading tiles from matrix $B$ into shared memory.
	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
	Performance is further improved to 614 GLOP/s after this optimisation.

@@

	AIWC also records the global memory address entropy (GMAE), a positive real number corresponding to the randomness of the memory access distribution of a program. To measure locality of memory accesses, AIWC collects the local memory address entropy (LMAE) of memory addresses accessed after dropping $n$ least significant bits of all memory addresses accessed by the program. To calculate this, AIWC collects a frequency distribution of all non-register memory accesses by all threads in the target kernel. Using the collected frequency distribution, AIWC calculates 10 separate local memory address entropy (LMAE) values according to increasing the number of least significant bits (LSB) skipped, using
	%in the following fashion. The information entropy of the collection of memory addresses accessed, after dropping $n$ LSBs and accounting for the number of memory accesses performed to each address is counted as the $n$-bits skipped LMAE. Given the frequency distribution of memory addresses accessed by the program, 
	the explicit formula for the $n$-bits skipped LMAE:

@@

	During simulation, AIWC collects a list of all memory accesses by each thread of execution.
	In the OpenCL programming model, threads within a work-group execute the same code and share access to \textit{local memory}.
	We can group together memory accesses of threads in a work-group at each logical time step in the symbolic execution of the code.
	On a GPU, memory accesses executed at the same time by different threads in a work-group are likely to interact, determining the extent of memory access coalescing and bank conflicts.

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Parallel spatial locality metric obtained from AIWC for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
		\label{fig: matmul parallel spatial locality}
	\end{figure*}

@@

	Needleman-Wunsch is a dynamic programming algorithm used to perform protein sequence alignment by identifying the similarity between two strings of amino acids.
	The computation of each element in the similarity matrix depends on its west, north and north-west neighbours.
	This dependency enforces a wave-front computation pattern, travelling along the main diagonal of the matrix. Each iteration of the kernels computes over an anti-diagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.

	In particular, the wave-front pattern of computations causes each work-item to request distant memory addresses in a parallel fashion. Thus, parallel memory requests into the input matrix prohibit locality. We note the lack of any drop-off in parallel spatial locality as the number of bits dropped increases. This rightly indicates that memory addresses accessed at each logical timestamp are very distant. On GPUs, this translates to poor utilization of both memory access coalescing and caching \cite{krommydas2016opendwarfs}. This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests -- this is typically performed in GPU optimized implementations of the Needleman-Wunsch algorithm \cite{opendwarfs2020head}.

@@

	Of the two BFS kernels, it is \texttt{BFS kernel 1} that performs the graph traversal necessary to generate a list of neighbours at each node. The memory access patterns in \texttt{BFS kernel 1} are irregular. 
	Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on. 
	The program structure involves multiple levels of pointer-chasing and thus the precise memory addresses accessed are data-dependent. 
	This leads to poor parallel spatial locality. Thus, we see a slow drop-off in parallel spatial locality for \texttt{BFS kernel 1}.

	\subsection{Sparse Linear Algebra: CSR}

@@

	In the OpenDwarfs implementation, each row of the input sparse matrix is assigned to a separate OpenCL work-item.
	The locations of matrix data read by each work-item are dependent on the number and position of non-zero values in the sparse-matrix, which are decided by the values in $Aj$ and $Ap$. 
	Similar to BFS, this means memory access patterns are runtime-dependent due to indirect addressing.
	This pattern of indirect addressing is typical of applications in the Spare Linear Algebra Dwarf, which severely hinders locality of memory accesses performed. The collected parallel spatial locality metric reflects this trend. Figure \ref{fig: opendwarfs parallel spatial locality} shows the gradual decline in parallel spatial locality as the number of bits dropped is increased.

	\section{Guidelines: Using the Parallel Spatial Locality Metric} \label{guidelines}

@@

	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as LUD and SRAD) will have a steep drop-off in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.

	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman–Wunsch, we observe a flat parallel spatial locality graph.
	 The worst possible memory access pattern: all threads in a work-group accessing results in a flat PSL metric of 1.
	 The PSL metric for Needleman–Wunsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate.
	 However, the workload imbalance across threads in each OpenCL work-group eases pressure on the memory unit when not all threads are active.

	 Codes with irregular or data-driven memory access patterns like CSR and BFS in the OpenDwarfs Suite have a slow drop-off in the PSL metric at low $n$ values. This is reflective of the mostly disparate parallel memory access pattern in these codes. The increasing steepness of the drop-off at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localisation of memory accesses.

>>>>

	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL. 
	%The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}. 
	This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.

@@

	The code is further optimized by improving the locality of reads from Matrix $B$.
	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
	The repeated reads of elements of matrix $B$ can be shared between threads in a work-group by reading tiles from matrix $B$ into shared memory.
	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
	Performance is further improved to 614 GLOP/s after this optimization.

@@

	AIWC also records the global memory address entropy (GMAE), a positive real number corresponding to the randomness of the memory access distribution of a program. To measure locality of memory accesses, AIWC collects the local memory address entropy (LMAE) of memory addresses accessed after dropping $n$ least significant bits of all memory addresses accessed by the program. To calculate this, AIWC collects a frequency distribution of all non-register memory accesses by all threads in the target kernel. Using the collected frequency distribution, AIWC calculates 10 separate local memory address entropy (LMAE) values according to increasing the number of least significant bits (LSB) skipped using
	%in the following fashion. The information entropy of the collection of memory addresses accessed, after dropping $n$ LSBs and accounting for the number of memory accesses performed to each address is counted as the $n$-bits skipped LMAE. Given the frequency distribution of memory addresses accessed by the program, 
	the explicit formula for the $n$-bits skipped LMAE:

@@

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Parallel spatial locality metric for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
		\label{fig: matmul parallel spatial locality}
	\end{figure*}

	During simulation, AIWC collects a list of all memory accesses by each thread of execution.
	In the OpenCL programming model, threads within a work-groups execute the same code and share access to \textit{local memory}.
	We can group together memory accesses of threads in a work-group at each logical time step in the symbolic execution of the code.
	On a GPU, memory accesses executed at the same time by different threads in a work-group are likely to interact, determining the extent of memory access coalescing and bank conflicts.

@@

	Needleman-Wunsch is a dynamic programming algorithm used to perform protein sequence alignment by identifying the similarity between two strings of amino acids.
	The computation of each element in the similarity matrix depends on its west, north and north-west neighbours.
	This dependency enforces a wave-front computation pattern, travelling along the main diagonal of the matrix. Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.

	In particular, the wavefront pattern of computations causes each work-item to request distant memory addresses in a parallel fashion. Thus, parallel memory requests into the input matrix prohibit locality. We note the lack of any drop-off in parallel spatial locality as the number of bits dropped increases. This rightly indicates that memory addresses accessed at each logical timestamp are very distant. On GPUs, this translates to poor utilization of both memory access coalescing and caching \cite{krommydas2016opendwarfs}. This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests -- this is typically performed in GPU optimized implementations of the Needleman-Wunsch algorithm \cite{opendwarfs2020head}.

@@

	Of the two BFS kernels, it is \texttt{BFS kernel 1} that performs the graph traversal necessary to generate a list of neighbours at each node. The memory access patterns in \texttt{BFS kernel 1} are irregular. 
	Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on. 
	The program structure involves multiple levels of pointer-chasing and thus the precise memory addresses accessed are data-dependent. 
	This leads to poor parallel spatial locality. Thus, we see a slow drop-off in parallel spatial locality for \texttt{BFS kernel 1}.
	\subsection{Sparse Linear Algebra: CSR}

@@

	In the OpenDwarfs implementation, each row of the input sparse matrix is assigned to a separate OpenCL work-item. 
	The locations of matrix data read by each work-item are dependent on the number and position of non-zero values in the sparse-matrix, which are decided by the values in $Aj$ and $Ap$. 
	Similar to BFS, this means memory access patterns are runtime-dependent due to indirect addressing.
	This pattern of indirect addressing is typical of applications in the Spare Linear Algebra Dwarf, which severely hinders locality of memory accesses performed. The collected parallel spatial locality metric reflects this trend. Figure \ref{fig: opendwarfs parallel spatial locality} shows the gradual decline in parallel spatial locality as the number of bits dropped is increased.
	\section{Guidelines: Using the Parallel Spatial Locality Metric} \label{guidelines}

@@

	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as in LUD and SRAD) will have a steep drop-off in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.

	 In the extreme case, when parallel memory accesses are very distant as for Needleman-Wunsch, we observe a flat parallel spatial locality graph. 
	 The worst possible memory access pattern occurs when all threads in a work-group access distant memory locations. This results in a flat PSL metric of 1. 
	 The PSL metric for Needleman-Wunsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate. 
	 However, the workload imbalance across threads in each OpenCL work-group eases pressure on the memory unit when not all threads are active.

	 Codes with irregular or data-driven memory access patterns like CSR and BFS in the OpenDwarfs Suite have a slow drop-off in the PSL metric at low $n$ values. This is reflective of the mostly disparate parallel memory access pattern in these codes. The increasing steepness of the drop-off at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localization of memory accesses.
